---
title: "Report Random Tapas"
date: "1/5/2021"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    
params:
  type: NULL
  phylo1: NULL
  mphylo1: NULL 
  phylo2: NULL
  mphylo2: NULL 
  relmat: NULL
  reps: NULL 
  dupli: NULL 
  n: NULL
  paco.ss: NULL 
  paco.warn: NULL
  paco.ei.corr: NULL 
  percentile: NULL
  link.res.fq: NULL 
  below.p: NULL
  colscale: NULL 
  n.breaks: NULL
  node.tag: NULL
  
bibliography: blibliography.bib
nocite: | 
  @Random, @PACo, @R, @GD
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE)
```

```{r librerias}
library(paco)
library(phytools)
library(distory)
library(GiniWegNeg)
library(parallel)
```

# Parameters

* The number of runs selected is: `r params$reps`
* The maximum number of n unique associations is: `r params$n`
* Assume symmetry in procrustes analysis: `r params$paco.ss`
* Correction used calculating the eigenvalues: `r params$paco.ei.corr`
* The percentile choosen is: `r params$percentile`
* Compute the Observed-Expected frequency: `r params$link.res.fq`
* Use only the associations below the percentile: `r params$below.p`




```{r funciones Random TaPas}
trimHS.maxC <- function (N, HS, n, check.unique= FALSE) {
trim.int <- function (x, HS, n) {
HS.LUT <- which(HS == 1, arr.in= TRUE)
HS.LUT <- cbind(HS.LUT, 1:nrow(HS.LUT))
df <- as.data.frame(HS.LUT)
hs.lut <- subset(df[sample(nrow(df)), ],
!duplicated(row) & !duplicated(col))
if (nrow(hs.lut) < n) hs <- NULL else {
hs.lut <- hs.lut[sample(nrow(hs.lut), n), ]
hs <- diag(nrow(hs.lut))
rownames(hs) <- rownames(HS[hs.lut[ ,1], ])
colnames(hs) <- colnames(HS[ ,hs.lut[ ,2]])
return(hs)
}
}
trim.HS <- lapply(1:N, trim.int, HS= HS, n= n )
if (check.unique == TRUE) trim.HS <- unique(trim.HS)
if (length(trim.HS) < N) warning("No. of trimmed H-S assoc. matrices < No. of runs")
return(trim.HS)
}

geo.D <- function (hs, treeH, treeS) {
treeh <- ape::drop.tip(treeH, setdiff(treeH$tip.label, rownames(hs)))
trees <- ape::drop.tip(treeS, setdiff(treeS$tip.label, colnames(hs)))
# foo distory requires same labels in both trees. Dummy labels are produced.
# 1st reorder hs as per tree labels:
hs <- hs[treeh$tip.label, trees$tip.label]
# 2nd swap trees labels with corresponding ones in treeh:
hs.lut <- which(hs[treeh$tip.label, trees$tip.label]==1, arr.ind = TRUE)
dummy.labels <- rownames(hs.lut)
trees$tip.label <- dummy.labels
combo.tree <- list(treeh, trees)
gd <- distory::dist.multiPhylo(combo.tree)
return(gd)
}

paco.ss <- function (hs, treeH, treeS, symmetric= FALSE,
proc.warns= FALSE, ei.correct= "none") {
eigen.choice <- c("none", "lingoes", "cailliez", "sqrt.D")
if (ei.correct %in% eigen.choice == FALSE)
stop(writeLines("Invalid eigenvalue correction parameter.\r
Correct choices are 'none', 'lingoes', 'cailliez' or 'sqrt.D'"))
treeh <- ape::drop.tip(treeH, setdiff(treeH$tip.label, rownames(hs)))
trees <- ape::drop.tip(treeS, setdiff(treeS$tip.label, colnames(hs)))
# Reorder hs as per tree labels:
hs <- hs[treeh$tip.label, trees$tip.label]
DH <- cophenetic(treeh)
DP <- cophenetic(trees)
if (ei.correct == "sqrt.D"){DH <- sqrt(DH) ; DP <- sqrt(DP); ei.correct="none"}
D <- paco::prepare_paco_data(DH, DP, hs)
D <- paco::add_pcoord(D, correction= ei.correct)
if (proc.warns == FALSE) D <- vegan::procrustes(D$H_PCo, D$P_PCo,
symmetric = symmetric) else
D <- suppressWarnings(vegan::procrustes(D$H_PCo, D$P_PCo,
symmetric = symmetric))
return(D$ss)
}

link.freq <- function (x, fx, HS, percentile= 0.01,
sep= "-", below.p= FALSE, res.fq= TRUE) {
if (below.p == TRUE)
percent <- which(fx <= quantile(fx, percentile)) else
percent <- which(fx >= quantile(fx, percentile))
trim.HS <- x[percent]
paste.link.names <- function(X, sep) {
X.bin <- which(X>0, arr.in=TRUE)
Y <- diag(nrow(X.bin))
Y <- diag(nrow(X.bin))
rownames(Y) <- rownames(X)[X.bin[,1]]
colnames(Y) <- colnames(X)[X.bin[,2]]
pln <- paste(rownames(Y), colnames(Y), sep=sep)
return(pln)
}
link.names <- t(sapply(trim.HS, paste.link.names, sep=sep))
lf <- as.data.frame(table(link.names))
HS.LUT <- which(HS ==1, arr.in=TRUE)
linkf <- as.data.frame(cbind(rownames(HS)[HS.LUT[,1]],
colnames(HS)[HS.LUT[,2]]))
colnames(linkf) <- c('H', 'S')
linkf$HS <- paste(linkf[,1], linkf[,2], sep=sep)
linkf$Freq <- rep(0, nrow(linkf))
linkf[match(lf[,1], linkf[,3]), 4] <- lf[,2]
linkf2 <- linkf
if (res.fq == TRUE) {
link.names.all <- t(sapply(x, paste.link.names, sep=sep))
lf.all <- as.data.frame(table(link.names.all))
linkf.all <- as.data.frame(cbind(rownames(HS)[HS.LUT[,1]],
colnames(HS)[HS.LUT[,2]]))
colnames(linkf.all) <- c('H', 'S')
linkf.all$HS <- paste(linkf.all[,1], linkf.all[,2], sep=sep)
linkf.all$Freq <- rep(0, nrow(linkf.all))
linkf.all[match(lf.all[,1], linkf.all[,3]), 4] <- lf.all[,2]
w <- linkf.all[,4]
w <- as.matrix(w*percentile)
wFq <- linkf$Freq - w
linkf$wFq <- wFq
} else linkf <- linkf2
return(linkf)
}

One2one.f <- function (hs, reps= 1e+4) {
HS.LUT <- which(hs ==1, arr.in=TRUE)
HS.LUT <- cbind(HS.LUT,1:nrow(HS.LUT))
df <- as.data.frame(HS.LUT)
V <- rep(NA,reps)
for(i in 1:reps){
hs.lut <- subset(df[sample(nrow(df)),],
!duplicated(row) & !duplicated(col))
n <- sum(HS)
while (n >0) {
n <- n-1;
if (nrow(hs.lut) == n) break
}
V[i]<- n
}
V <- min(V)
return(V)
}

tangle.gram <- function(treeH, treeS, hs, colscale= "diverging", colgrad,
nbreaks=50, fqtab, res.fq= TRUE, node.tag=TRUE,
cexpt=1, ...) {
colscale.choice <- c("diverging", "sequential")
if (colscale %in% colscale.choice == FALSE)
stop(writeLines("Invalid colscale parameter.\r
Correct choices are 'diverging', 'sequential'"))
colscale.range <- function(x) {
rescale.range <- function(x) {
xsq <- round(x)
if(colscale=="sequential") {
y <- range(xsq)
col_lim <- (y[1]:y[2])-y[1]+1
xsq <- xsq-y[1]+1
new.range <- list(col_lim, xsq)
} else {
x1 <- x[which(x<0)]
if(length(x1) < 2) stop("Not enough negative values for diverging scale.
Choose colscale= 'sequential' instead")
x2 <- x[which(x >= 0)]
x1 <- round(x1)
x2 <- round(x2)
y <- max(abs(x))
col_lim <- (-y:y) + y + 1
y1 <- range(x1)
y2 <- range(x2)
x1 <- x1-y1[1]+1
x2 <- x2-y2[1]+1
new.range <- list(col_lim, x1, x2)
}
return(new.range)
}
if(colscale=="sequential") {
NR <- rescale.range(x)
rbPal <- colorRampPalette(colgrad)
linkcolor <- rbPal(nbreaks)[as.numeric(cut(NR[[1]], breaks = nbreaks))]
NR <- NR[[2]]
linkcolor <- linkcolor[NR]
} else {
NR <- rescale.range(x)
NR.neg <- NR[[1]] [which (NR[[1]] <= max(NR[[2]]))]
NR.pos <- NR[[1]] [-NR.neg] - max(NR[[2]])
m <- median(1:length(colgrad))
colgrad.neg <- colgrad[which(1:length(colgrad) <= m)]
colgrad.pos <- colgrad[which(1:length(colgrad) >= m)]
rbPal <- colorRampPalette(colgrad.neg)
linkcolor1 <- rbPal(nbreaks)[as.numeric(cut(NR.neg, breaks = nbreaks))]
rbPal <- colorRampPalette(colgrad.pos)
linkcolor2 <- rbPal(nbreaks)[as.numeric(cut(NR.pos, breaks = nbreaks))]
linkcolor1 <- linkcolor1[NR[[2]]]
linkcolor2 <- linkcolor2[NR[[3]]]
linkcolor <- rep(NA, length(x))
linkcolor[which(x< 0)] <- linkcolor1
linkcolor[which(x>=0)] <- linkcolor2
}
return(linkcolor)
}
FQ <- ifelse(res.fq==FALSE, 4 ,5) # determines freq column to evaluate
if (res.fq ==FALSE & colscale== "diverging") {colscale = "sequential"
warning("Colscale 'diverging' does not take effect when res.fq = FALSE.
The color scale shown is sequential")
}
LKcolor <- colscale.range(fqtab[,FQ])
HS.lut <- which(hs ==1, arr.ind=TRUE)
linkhs <- cbind(rownames(hs)[HS.lut[,1]], colnames(hs)[HS.lut[,2]])
obj <- phytools::cophylo(treeH,treeS, linkhs)
phytools::plot.cophylo(obj, link.col=LKcolor, ...)
Hfreq <- aggregate(fqtab[,FQ], by=list(freq = fqtab[,1]), FUN=mean)
Sfreq <- aggregate(fqtab[,FQ], by=list(freq = fqtab[,2]), FUN=mean)
Hfreq <- Hfreq[match(obj$trees[[1]]$tip.label, Hfreq$freq),]
Sfreq <- Sfreq[match(obj$trees[[2]]$tip.label, Sfreq$freq),]
Hfreq$x[which(is.na(Hfreq$x))]<-0
Sfreq$x[which(is.na(Sfreq$x))]<-0
if (node.tag==TRUE){
fit.H <- phytools::fastAnc(obj$trees[[1]],Hfreq[,2])
fit.S <- phytools::fastAnc(obj$trees[[2]],Sfreq[,2])
NLH <- colscale.range (fit.H)
NLS <- colscale.range (fit.S)
phytools::nodelabels.cophylo(pch=16, col=NLH, cex=cexpt)
phytools::nodelabels.cophylo(pch=16, col=NLS, cex=cexpt, which="right")
}
TLH <- colscale.range (Hfreq[,2])
TLS <- colscale.range (Sfreq[,2])
phytools::tiplabels.cophylo(pch=16, col=TLH, cex=cexpt)
phytools::tiplabels.cophylo(pch=16, col=TLS, cex=cexpt, which="right")
}


```

```{r, reading data}
if(params$type=="Nexus"){
  TreeH<-read.nexus(params$phylo1)
  TreeS<-read.nexus(params$phylo2)
  mTreeH<-read.nexus(params$mphylo1)
  mTreeS<-read.nexus(params$mphylo2)
  HS<-read.table(params$relmat, row.names = 1, header = T)
  HS<-as.matrix(HS)
}

if(params$type=="Newick"){
  TreeH<-read.tree(params$phylo1)
  TreeS<-read.tree(params$phylo2)
  mTreeH<-read.tree(params$mphylo1)
  mTreeS<-read.tree(params$mphylo2)
  HS<-read.table(params$relmat, row.names = 1, header = T)
  HS<-as.matrix(HS)
}
```

# Tanglegram

Plain tanglegram representing the associations between the two
phyllogenies. The number of species of the host phyllogeny is: `r length(TreeH$tip.label)`,
the number of species of the symbiont phyllogeny is `r length(TreeS$tip.label)`.
The number of associations between the two phyllogenies is `r sum(HS)`.

```{r, plain tanglegram, fig.cap="Tanglegram showing the associations betweewn the two philogenies"}
HS.lut <- which(HS ==1, arr.ind=TRUE)
linkhs <- cbind(rownames(HS)[HS.lut[,1]], colnames(HS)[HS.lut[,2]])
obj <- cophylo(TreeH,TreeS, linkhs)
plot.cophylo(obj, link.lwd=1, link.lty=2, fsize=0.5, pts=FALSE, link.type="curved")
```

```{r, plotting the optimal number of n associations for Random TaPas}
N = params$reps
```

# Extracting the partial tanglegrams

A total of `r length(THS)` different partial tanglegrams have been extracted 
from the cophylogeny.

```{r, fig.cap="Four random tanglegrams computed by the function trimHS.maxC"}
par(mfrow = c(2,2), oma = rep(2,4))

randomtree<-sample(1:length(THS), replace = F, size = 4)

for (i in randomtree) {
  hs<-THS[[i]]
  HS.lut <- which(hs ==1, arr.ind=TRUE)
  linkhs <- cbind(rownames(hs)[HS.lut[,1]], colnames(hs)[HS.lut[,2]])
  treeh <- ape::drop.tip(TreeH, setdiff(TreeH$tip.label, rownames(hs)))
  trees <- ape::drop.tip(TreeS, setdiff(TreeS$tip.label, colnames(hs)))
  obj<-cophylo(tr1 =  treeh,tr2 =  trees,assoc =  linkhs)
  plot.cophylo(obj,link.type = "curved")
}
```

# Distribution of the statistics

```{r, computing the geodesic distances and the procrustes sum square residuals, fig.cap="Distribution of the Geodesic distances and SSR of PACo analyses"}
n = params$n
if(params$paco.ss == TRUE){
  sim<-TRUE
}else{
  sim<-FALSE
}

perc<-as.numeric(params$percentile)

THS <- trimHS.maxC(N, HS, n=n, check.unique=params$dupli)
THS[sapply(THS, is.null)] <- NULL
GD <- sapply(THS, geo.D, treeH=TreeH, treeS= TreeS)
PACO <- sapply(THS, paco.ss, treeH=TreeH, treeS= TreeS, ei.correct = params$paco.ei.corr,
symmetric = sim)

LFGD01 <- link.freq(THS, GD, HS, percentile=perc)
LFPACO01 <- link.freq(THS, PACO, HS, percentile=perc)

GD01 <- matrix(NA, length(mTreeH), nrow(LFGD01))
PACO01 <- matrix(NA, length(mTreeH), nrow(LFPACO01))

par(mfrow = c(2,1))
densGD<-density(GD)
quanGD<-quantile(GD, probs = 0.01)

densPACO<-density(PACO)
quanPACO<-quantile(PACO, probs = 0.01)

plot(densGD, main = "Density plot of the distribution of the Geodesic Distances",
     xlab = "Geodesic distance")
abline(v=quanGD, col = "red")
plot(densPACO, main = "Density plot of the distribution of the SSR of PACo",
     xlab = "sum of squared residuals of PACo")
abline(v=quanGD, col = "red")
par(mfrow = c(1,1))
```

```{r, Average residual frequency distributions}
cores <- detectCores()
cl <- makeCluster(cores-2)
# use no CPUs - 2 for parallel computing
for(i in 1:length(mTreeH))
# CIs GD
{
GD.CI<-parallel::parSapply(cl, THS, geo.D, treeH=mTreeH[[i]],
treeS= mTreeS[[i]])
LFGD01.CI <- link.freq(THS, GD.CI, HS, percentile=perc)
GD01[i,] <- LFGD01.CI[,5]
}
#
for(i in 1:length(mTreeH))
# CIs PACo
{
PA.CI<-parallel::parSapply(cl, THS, paco.ss, treeH=mTreeH[[i]],
treeS= mTreeS[[i]], symmetric=sim)
LFPA01.CI <- link.freq(THS, PA.CI, HS, percentile=perc)
PACO01[i,] <- LFPA01.CI[,5]
}
stopCluster(cl)
#
colnames(GD01) <- LFGD01[,3]
colnames(PACO01) <- LFPACO01[,3]

GD.LO <- apply(GD01, 2, quantile, 0.025)
GD.HI <- apply(GD01, 2, quantile, 0.975)
GD.AV <- apply(GD01, 2, mean)
#
PACO.LO <- apply(PACO01, 2, quantile, 0.025)
PACO.HI <- apply(PACO01, 2, quantile, 0.975)
PACO.AV <- apply(PACO01, 2, mean)

cap4<-sprintf("Average residual frequency distributions (observed – expected frequencies) of host-symbiont applying Random TaPas with Geodesic distances and PACo , N = %s, n = %.0f and p = %.2f. Vertical lines represent 95 percent confidence intervals of the residual frequencies
computed empirically with %.0f randomly chosen pairs of trees used to build the consensus trees", N, n, perc, length(mTreeH))
```

```{r, plotting OE frequencies, fig.cap=cap4}
op <- par(mfrow=c(2,1),mgp = c(2.2, 0.3, 0), mar=c(3.5,3.2,1,0), tck=0.01,
oma = c(5.4, 0, 0, 0), xpd=NA)
link.fq <-barplot(GD.AV, xaxt='n',
horiz=FALSE, cex.names = 0.6, las=2, cex.axis=0.8,
ylab="Observed - Expected frequency", main = "GD",
ylim=c(min(GD.LO), max(GD.HI)), col="lightblue")
suppressWarnings(arrows(link.fq, GD.HI, link.fq, GD.LO, length= 0,
angle=90, code=3, col="darkblue"))
link.fq <-barplot(PACO.AV, xaxt='n',
horiz=FALSE, cex.names = 0.6, las=2, cex.axis=0.8,
ylab="Observed - Expected frequency", main = "PACo",
ylim=c(min(GD.LO), max(PACO.HI)),col="lightblue")
suppressWarnings(arrows(link.fq, PACO.HI, link.fq, PACO.LO, length= 0,
angle=90, code=3, col="darkblue"))
axis(side=1, at=link.fq[1:length(PACO.AV)], labels=LFPACO01$HS, las=2,
tick = FALSE, line= 5, cex.axis=0.6)
par(op)

cap5<-sprintf("Normalized Gini coefficient of the residual frequency distributions applying Random TaPas with geodesic distances and PACo to the consensus trees (asterisk) and to %.0f randomly chosen trees used to build the consensus trees", length(mTreeH))
```

# Normalized Gini coefficients

```{r, plotting Ginnis coef, fig.cap = cap5}
GiniGD <- unlist(Gini_RSV(LFGD01[,5]))
GiniPA <- unlist(Gini_RSV(LFPACO01[,5]))
GiniMGD <- unlist(apply(GD01, 1, Gini_RSV))
GiniMPA <- unlist(apply(PACO01, 1, Gini_RSV))
boxplot(GiniMGD, GiniMPA, names = c("GD", "PACo"), ylab="Normalized Gini coefficient",
col="lightblue", las=3)
text(1,GiniGD,"*",cex=2, col="red")
text(2,GiniPA,"*",cex=2, col="red")
```

* The average Gini coefficient usig the geodesic distances is `r GiniGD`.
* The average Gini coefficient usig the geodesic distances is `r GiniPA`


# Heatmap tanglegram with the Geodesic distance estimations

```{r, plotting tanglegram heatmap, fig.cap = "Tanglegram representing the association between hosts and symbionts. The residual frequency computed using the geodesic distances corresponding to each host-symbiont association is mapped using a color scale centered at light gray (zero) ranging from dark red (lowest) to dark blue (highest). The average residual frequency of occurrence of each terminal and fast maximum likelihood estimators of ancestral states of each node are also mapped according to the same scale."}
trh <- compute.brtime(TreeH, TreeH$Nnode:1)
trs <- compute.brtime(TreeS, TreeS$Nnode:1)
col.scale <- c("darkred","gray90", "darkblue")
tangle.gram(trh, trs, HS, colscale= params$colscale, colgrad=col.scale,
nbreaks=as.numeric(params$n.breaks), LFGD01, res.fq=params$link.res.fq, link.lwd=1, link.lty=1, fsize=0.5,
pts=FALSE, link.type="curved", node.tag=params$node.tag,
cexpt=1.2, ftype="reg")
```

# Heatmap tanglegram with the PACo estimations

```{r, plotting tanglegram heatmap, fig.cap = "Tanglegram representing the association between hosts and symbionts. The residual frequency computed using PACo corresponding to each host-symbiont association is mapped using a color scale centered at light gray (zero) ranging from dark red (lowest) to dark blue (highest). The average residual frequency of occurrence of each terminal and fast maximum likelihood estimators of ancestral states of each node are also mapped according to the same scale."}
trh <- compute.brtime(TreeH, TreeH$Nnode:1)
trs <- compute.brtime(TreeS, TreeS$Nnode:1)
col.scale <- c("darkred","gray90", "darkblue")
tangle.gram(trh, trs, HS, colscale= params$colscale, colgrad=col.scale,
nbreaks=as.numeric(params$n.breaks), LFPACO01, res.fq=params$link.res.fq, link.lwd=1, link.lty=1, fsize=0.5,
pts=FALSE, link.type="curved", node.tag=params$node.tag,
cexpt=1.2, ftype="reg")
```

# References